# Installation {#installation}

## Requirements {-}

* **R ≥ 4.x** with hard dependencies: `hdf5r`, `Matrix`, `methods`,
  `data.table`, `R.utils`.
* **Python ≥ 3.9** with hard dependencies: `shiny`, `shinywidgets`,
  `h5py`, `numpy`, `scipy`, `pandas`, `plotly`.

Optional R Suggests (needed only when you exercise certain code paths):

* `Biobase` (Bioconductor) — for ExpressionSet inputs to
  `prepare_study()` / `prepare_study_from_eset()`.
* `yaml` — for `prepare_study()`'s YAML config entry point.
* `shiny`, `bslib`, `plotly`, `DT`, `visNetwork`, `htmlwidgets` — for
  the Shiny app.
* `roxygen2` — only needed if you edit source comments and want to
  regenerate man pages.

## R / RStudio {-}

### From the shell {-}

```sh
R CMD build scminerViewer
R CMD INSTALL scminerViewer_0.1.0.tar.gz
```

### From R or RStudio {-}

```{r install-r, eval=FALSE}
# devtools / pak / remotes all work — choose whichever you already have
install.packages("devtools")
devtools::install("scminerViewer")     # run from the project root
```

In **RStudio**:

1. **File → Open Project…** and pick `scminerViewer/` (it ships a
   `.Rproj`).
2. **Build → Install Package** (`Cmd/Ctrl + Shift + B`).
3. To install all Suggests too:

   ```{r install-suggests, eval=FALSE}
   install.packages(c("yaml", "shiny", "bslib", "plotly", "DT",
                       "visNetwork", "htmlwidgets"))
   if (!requireNamespace("BiocManager", quietly = TRUE))
     install.packages("BiocManager")
   BiocManager::install("Biobase")
   ```

### Verify {-}

```{r verify-r, eval=FALSE}
library(scminerViewer)
?prepare_study      # opens the man page
```

If `?prepare_study` returns "No documentation found", the man pages
weren't generated. From the package source:

```{r roxy, eval=FALSE}
roxygen2::roxygenise("scminerViewer")  # writes NAMESPACE + man/*.Rd
```

then reinstall.

## Python {-}

```sh
python3 -m venv .venv
source .venv/bin/activate
pip install -e "scminer_viewer[dev]"
```

This installs the runtime deps plus `pytest` / `pytest-cov`. The CLI
is available as `scminer-viewer`:

```sh
scminer-viewer --help
```

Verify:

```sh
scminer-viewer list data
scminer-viewer info data/2327/2327.scminer.h5
```

## Common installation issues {-}

### "lazy-load database is corrupt" / "interrupted promise evaluation"

R memory-maps the package's `.rdb` file during `library()`. If a second
R process reinstalls the package while the first one still has it
loaded, the offsets go stale and you'll see one of:

```
Error: lazy-load database '.../scminerViewer.rdb' is corrupt
Warning: restarting interrupted promise evaluation
Error in R_decompress1
```

The cure is always:

1. **Quit all R sessions** that have the package loaded (`Cmd/Ctrl + Q`
   in RStudio; this exits the R process — restarting alone isn't
   enough).
2. **Wipe the install directory** from the shell:

   ```sh
   rm -rf "$HOME/StJude/Research/R_lib/scMINERViewer/scminerViewer"
   ```
3. **Reinstall from a fresh R session**:

   ```{r reinstall, eval=FALSE}
   .libPaths(c("$HOME/StJude/Research/R_lib/scMINERViewer", .libPaths()))
   devtools::install("scminerViewer", upgrade = "never")
   ```

### Missing system libraries on macOS

`hdf5r` needs HDF5; if `install.packages("hdf5r")` fails, install
HDF5 with Homebrew first:

```sh
brew install hdf5
```

then retry the R install.
