# Tutorial: from zero to a multi-study browser {#tutorial}

This chapter walks through the most common workflow end-to-end:
prepare one study, browse it, add a second study, browse them
together. All paths assume the project root as the working directory.

## A — Prepare one study from a YAML config {-}

```{r tutorial-yaml, eval=FALSE}
# 1) Copy the annotated template that ships with the package
file.copy(
  system.file("extdata", "example_config.yml", package = "scminerViewer"),
  "config.yml"
)
```

Open `config.yml` and edit:

* `output:` — the **root** that will host all your studies (e.g. `"data"`).
* `study.*` — ID, abbreviation, titles for this study.
* `input.expression` — path to your expression `ExpressionSet` RDS.
* `input.activity` (optional) — path to your activity ExpressionSet.
* `input.networks` (optional) — path to the networks TSV.

Then:

```{r tutorial-prepare, eval=FALSE}
library(scminerViewer)
prepare_study("config.yml")
```

This writes `data/<studyID>/<studyID>.scminer.h5` plus the full graph
layout (Cell/, Gene/, Network_*/, …) and the per-gene shard tree
(`expression_files/<studyID>/<letter>/<gene>.csv.gz` and
`activity_files/<studyID>/{TF,SIG}/...`) into `data/<studyID>/`.

## B — Migrate an existing on-disk study (no RDS needed) {-}

If you already have the graph-layout TSVs + shard tree on disk
(e.g. under `data/example/`), build the bundle directly without
invoking the Biobase pipeline:

```{r tutorial-migrate, eval=FALSE}
study <- read_graph_study(data_dir = "data/example", study_id = "2327")
dir.create("data/2327", recursive = TRUE, showWarnings = FALSE)
write_bundle(
  bundle_path        = "data/2327/2327.scminer.h5",
  meta               = study$meta,
  cells              = study$cells,
  clusters           = study$clusters,
  genes              = study$genes,
  expression_genes   = study$expression_genes,
  activity_tf_genes  = study$activity_tf_genes,
  activity_sig_genes = study$activity_sig_genes,
  network_tf         = study$network_tf,
  network_sig        = study$network_sig,
  overwrite          = TRUE
)
```

If you also want the shard tree co-located with the new bundle, move
it once:

```sh
mv data/example/expression_files data/2327/
mv data/example/activity_files   data/2327/
```

Otherwise pass `shard_dir =` to `run_app()` / `run_browser()` (see C).

## C — Browse one or many studies {-}

### Single study (R) {-}

```{r tutorial-runapp, eval=FALSE}
# Bundle and shards co-located (default — auto-discovered)
run_app("data/2327/2327.scminer.h5", port = 8000)

# Shards live elsewhere
run_app("data/2327/2327.scminer.h5",
        shard_dir = "data/example", port = 8000)
```

### Single study (Python) {-}

```sh
scminer-viewer run data/2327/2327.scminer.h5 --port 8000
scminer-viewer run data/2327/2327.scminer.h5 \
                   --shard-dir data/example --port 8000
```

### Multi-study browser (R) {-}

```{r tutorial-browser-r, eval=FALSE}
run_browser("data", port = 8000)

# When shards still live in the legacy data/example/ tree
run_browser("data", shard_dir = "data/example", port = 8000)
```

### Multi-study browser (Python) {-}

```sh
scminer-viewer browse data --port 8000
scminer-viewer browse data --shard-dir data/example --port 8000
```

Inside the browser, each study is a card; clicking opens
`?study=<sid>` (R) or sets a reactive value (Python) to drill in.
A "← Back to studies" link returns to the index.

## D — Adding a second study {-}

Point `prepare_study()` at the same root with a new `study.ID`. Each
call creates a sibling `<sid>/` folder:

```
data/
├── 2327/                      # first study
│   └── 2327.scminer.h5
└── 9999/                      # second study
    └── 9999.scminer.h5
```

`run_browser("data")` (or `scminer-viewer browse data`) now shows two
cards. No rebuild of the existing study is required.

## E — Inspect what's available without launching anything {-}

`discover_studies()` returns one row per bundle found under a root:

```{r tutorial-discover-r, eval=FALSE}
scminerViewer::discover_studies("data")
#    studyID studyAbbr shortTitle              longTitle    species n_cells n_genes n_clusters bundle_path                study_dir
# 1     2327       tex        Tex Single-cell RNA-seq … Mus musculus    8464    9861          3 data/2327/2327.scminer.h5 data/2327
```

```sh
scminer-viewer list data
# studyID studyAbbr shortTitle      species  n_cells  n_genes  n_clusters  bundle_path
#    2327       tex        Tex Mus musculus     8464     9861           3  data/2327/2327.scminer.h5
```

Useful for sanity-checking your layout before launching a browser
session, or for scripting (e.g. iterate over every study's bundle in
a batch job).
