# Tutorial: from zero to a multi-study browser {#tutorial}

This chapter walks through the most common workflow end-to-end:
prepare one study, browse it, add a second study, browse them
together. All paths assume the project root as the working directory.

## A — Walkthrough with the public 2327 (Tex) demo data {-}

If you don't yet have your own scMINER outputs, the project ships a
complete worked example: the 2327 CD8⁺ T-cell exhaustion study from
*"Single-cell RNA sequencing of progenitor, intermediate and
terminally exhausted CD8+ T cells from tumors and chronic viral
infection"*. The three source files are hosted as a GitHub Release
asset on this repo.

> **Don't want to run `prepare_study()`?** Skip to
> [A.2 — Fast path](#tutorial-2327-fast-path) and download the
> pre-processed bundle directly.

### A.1 — Run `prepare_study()` from scratch {-}

#### 1. Download the inputs from the release {-}

```sh
mkdir -p data/input/2327 && cd data/input/2327

RELEASE_URL=https://github.com/hzhou98/scMINER-Viewer/releases/download/demo-data-v1

curl -L -o expression.rds "$RELEASE_URL/expression.rds"   #  22 MB
curl -L -o activity.rds   "$RELEASE_URL/activity.rds"     # 236 MB
curl -L -o networks.txt   "$RELEASE_URL/networks.txt"     #  57 MB
```

Verify the downloads (optional but recommended for the 236 MB file):

```sh
shasum -a 256 expression.rds activity.rds networks.txt
# expression.rds  9c28047387a181263107cab3076d426aaa32fcc402e6a2af6b4fb1ec5b910960
# activity.rds    10b94c7826ce513c792ce0cdd57f6c1cad1adb3576f8f6f17779d35b6413c487
# networks.txt    9096a1c3604a93e19e2ac177728686ca0f45c6de48a9b27149fc6e6e9b8dc856
```

About the inputs:

* `expression.rds` — Biobase `ExpressionSet` (genes × cells), normalised
  counts. `pData` carries `CellID`, `CellGroup`, and `UMAP_1` / `UMAP_2`
  coordinates; `fData` carries `GeneSymbol`.
* `activity.rds` — Biobase `ExpressionSet` with the same cells. Row names
  end in `_TF` or `_SIG`; `prepare_study()` splits them into the
  TF-activity and SIG-activity matrices automatically.
* `networks.txt` — tab-separated; columns include `source`, `target`,
  `MI`, `pearson`, `spearman`, `slope`, `p.value`, `CellGroup`, and
  `NetworkType` ∈ {TF, SIG}.

#### 2. Write a YAML config that points at them {-}

```yaml
# config-2327.yml
output: data        # writes data/2327/2327.scminer.h5 + the shard tree

study:
  ID:         "2327"
  studyAbbr:  Tex
  shortTitle: Tex
  longTitle:  >
    Single-cell RNA sequencing of progenitor, intermediate and
    terminally exhausted CD8+ T cells from tumors and chronic
    viral infection

species:    Mus musculus
coordinate: UMAP

cellID:     CellID
cellType:   CellGroup
cellGroup:  CellGroup
geneSymbol: GeneSymbol

input:
  expression: data/input/2327/expression.rds
  activity:   data/input/2327/activity.rds
  networks:   data/input/2327/networks.txt

cluster_palette: npg
default_genes: [Cd8a, Pdcd1, Tox]
```

`prepare_study()` resolves `input.*` paths relative to the current
working directory — keep your shell in the project root when you run
it, or rewrite the paths to be absolute.

#### 3. Run `prepare_study()` {-}

```{r tutorial-2327-prepare, eval=FALSE}
library(scminerViewer)
prepare_study("config-2327.yml")
```

Expect 1–3 minutes on a laptop (most of the time goes to writing the
~15,500 per-gene shards). When it finishes, you'll have:

```
data/
└── 2327/
    ├── 2327.scminer.h5                # bundle (~77 MB)
    ├── Cell/  Gene/  Network_*/  …    # graph layout
    ├── expression_files/2327/{meta.csv, <letter>/<gene>.csv.gz}
    └── activity_files/2327/{meta.csv, {TF,SIG}/<letter>/<gene>.csv.gz}
```

#### 4. Launch the viewer {-}

```{r tutorial-2327-runapp, eval=FALSE}
run_app("data/2327/2327.scminer.h5", port = 8000)
```

The gene picker pre-selects `Cd8a, Pdcd1, Tox` thanks to the
`default_genes:` line in the YAML, so you land on a populated view.

### A.2 — Fast path: download the pre-processed bundle {#tutorial-2327-fast-path -}

If you just want to *see* the viewer without running `prepare_study()`,
download the already-prepared bundle + shard tree as a single tarball:

```sh
mkdir -p data && cd data

RELEASE_URL=https://github.com/hzhou98/scMINER-Viewer/releases/download/demo-data-v1
curl -L -o 2327-processed.tar.gz "$RELEASE_URL/2327-processed.tar.gz"  # ~484 MB

shasum -a 256 2327-processed.tar.gz
# 2327-processed.tar.gz  c7e361c5623be787ca2874d4ef937a9bc38a59a44e3ee5b0ac7415ec656d7680

tar -xzf 2327-processed.tar.gz
mv 2327-processed 2327
```

Then:

```{r tutorial-2327-fastpath, eval=FALSE}
scminerViewer::run_app("data/2327/2327.scminer.h5", port = 8000)
```

> The R package also ships an even smaller in-package demo (200 gene
> shards, ~80 MB total) accessible via `run_demo()` — see
> [Installation](#installation). Use it when you want the very fastest
> "what does this look like" path with zero downloads beyond the package
> itself.

## B — Prepare one study from a YAML config {-}

```{r tutorial-yaml, eval=FALSE}
# 1) Copy the annotated template that ships with the package
file.copy(
  system.file("extdata", "example_config.yml", package = "scminerViewer"),
  "config.yml"
)
```

Open `config.yml` and edit:

* `output:` — the **root** that will host all your studies (e.g. `"data"`).
* `study.*` — ID, abbreviation, titles for this study.
* `input.expression` — path to your expression `ExpressionSet` RDS.
* `input.activity` (optional) — path to your activity ExpressionSet.
* `input.networks` (optional) — path to the networks TSV.

Then:

```{r tutorial-prepare, eval=FALSE}
library(scminerViewer)
prepare_study("config.yml")
```

This writes `data/<studyID>/<studyID>.scminer.h5` plus the full graph
layout (Cell/, Gene/, Network_*/, …) and the per-gene shard tree
(`expression_files/<studyID>/<letter>/<gene>.csv.gz` and
`activity_files/<studyID>/{TF,SIG}/...`) into `data/<studyID>/`.

## C — Migrate an existing on-disk study (no RDS needed) {-}

If you already have the graph-layout TSVs + shard tree on disk
(e.g. under `data/example/`), build the bundle directly without
invoking the Biobase pipeline:

```{r tutorial-migrate, eval=FALSE}
study <- read_graph_study(data_dir = "data/example", study_id = "2327")
dir.create("data/2327", recursive = TRUE, showWarnings = FALSE)
write_bundle(
  bundle_path        = "data/2327/2327.scminer.h5",
  meta               = study$meta,
  cells              = study$cells,
  clusters           = study$clusters,
  genes              = study$genes,
  expression_genes   = study$expression_genes,
  activity_tf_genes  = study$activity_tf_genes,
  activity_sig_genes = study$activity_sig_genes,
  network_tf         = study$network_tf,
  network_sig        = study$network_sig,
  overwrite          = TRUE
)
```

If you also want the shard tree co-located with the new bundle, move
it once:

```sh
mv data/example/expression_files data/2327/
mv data/example/activity_files   data/2327/
```

Otherwise pass `shard_dir =` to `run_app()` / `run_browser()` (see D).

## D — Browse one or many studies {-}

### Single study (R) {-}

```{r tutorial-runapp, eval=FALSE}
# Bundle and shards co-located (default — auto-discovered)
run_app("data/2327/2327.scminer.h5", port = 8000)

# Shards live elsewhere
run_app("data/2327/2327.scminer.h5",
        shard_dir = "data/example", port = 8000)
```

### Single study (Python) {-}

```sh
scminer-viewer run data/2327/2327.scminer.h5 --port 8000
scminer-viewer run data/2327/2327.scminer.h5 \
                   --shard-dir data/example --port 8000
```

### Multi-study browser (R) {-}

```{r tutorial-browser-r, eval=FALSE}
run_browser("data", port = 8000)

# When shards still live in the legacy data/example/ tree
run_browser("data", shard_dir = "data/example", port = 8000)
```

### Multi-study browser (Python) {-}

```sh
scminer-viewer browse data --port 8000
scminer-viewer browse data --shard-dir data/example --port 8000
```

Inside the browser, each study is a card; clicking opens
`?study=<sid>` (R) or sets a reactive value (Python) to drill in.
A "← Back to studies" link returns to the index.

## E — Adding a second study {-}

Point `prepare_study()` at the same root with a new `study.ID`. Each
call creates a sibling `<sid>/` folder:

```
data/
├── 2327/                      # first study
│   └── 2327.scminer.h5
└── 9999/                      # second study
    └── 9999.scminer.h5
```

`run_browser("data")` (or `scminer-viewer browse data`) now shows two
cards. No rebuild of the existing study is required.

## F — Inspect what's available without launching anything {-}

`discover_studies()` returns one row per bundle found under a root:

```{r tutorial-discover-r, eval=FALSE}
scminerViewer::discover_studies("data")
#    studyID studyAbbr shortTitle              longTitle    species n_cells n_genes n_clusters bundle_path                study_dir
# 1     2327       tex        Tex Single-cell RNA-seq … Mus musculus    8464    9861          3 data/2327/2327.scminer.h5 data/2327
```

```sh
scminer-viewer list data
# studyID studyAbbr shortTitle      species  n_cells  n_genes  n_clusters  bundle_path
#    2327       tex        Tex Mus musculus     8464     9861           3  data/2327/2327.scminer.h5
```

Useful for sanity-checking your layout before launching a browser
session, or for scripting (e.g. iterate over every study's bundle in
a batch job).
