# Config YAML reference {#config-yaml}

`prepare_study(config_path)` is the YAML-driven entry point for the
full pipeline. It points at a scMINER `ExpressionSet` RDS (and
optional activity + networks) and writes both the graph-import layout
and the lazy-mode bundle in one call.

Two examples ship with the project:

| File | Role |
| --- | --- |
| `scminerViewer/inst/extdata/example_config.yml` | Annotated **template** — every field documented inline; reachable via `system.file()`. |
| `data/example/2327.yml`                          | Concrete **runnable** config keyed to the actual 2327 study (Tex). Edit `input.*` paths and run. |

Copy the template:

```{r copy-template, eval=FALSE}
file.copy(
  system.file("extdata", "example_config.yml", package = "scminerViewer"),
  "config.yml"
)
```

## Full schema {-}

```yaml
# Where to write everything — the bundle lands at
# <output>/<studyID>/<studyID>.scminer.h5 (per-study subfolder).
output: data

# Study identity
study:
  ID:         "2327"
  studyAbbr:  tex
  longTitle:  >
    Single-cell RNA sequencing of progenitor, intermediate and
    terminally exhausted CD8+ T cells from tumors and chronic viral
    infection
  shortTitle: Tex

# Sample-level metadata
species:    Mus musculus
coordinate: UMAP            # also the pData column prefix (UMAP_1, UMAP_2)

# Cluster colour palette — auto-fills the study_meta.csv `Color` column
# and the bundle's clusters/color group. Pass any ggsci palette name:
#   npg, aaas, lancet, nejm, jama, jco, ucscgb, d3, locuszoom, igv,
#   uchicago, startrek, tron, futurama, rickandmorty, simpsons
# See <https://cran.r-project.org/web/packages/ggsci/vignettes/ggsci.html>.
cluster_palette: npg        # default — Nature Publishing Group

# pData / fData column names in the source ExpressionSet
cellID:     cellID          # rownames are also assigned to this column
cellType:   cellGroup       # the cell-type / cluster label
cellGroup:  cellGroup       # falls back to cellType when omitted
geneSymbol: geneSymbol      # column in fData carrying the gene symbol

# Input files
input:
  expression: inputs/2327_expression.rds   # required, Biobase ExpressionSet
  activity:   inputs/2327_activity.rds     # optional, ExpressionSet
                                            # rows ending in _TF / _SIG split
                                            # into the two activity matrices
  networks:   inputs/2327_networks.tsv     # optional, tab-separated:
                                            # source, target, NetworkType,
                                            # CellGroup, mi, pearson,
                                            # spearman, rho, pvalue

# Optional: genes the Shiny app loads on launch. Mirrors the scMINER
# Portal's preGenes field. Any of these four shapes parse to the same
# normalised character vector internally; pick whichever reads
# cleanest in your YAML:
default_genes: [Cd36, Pdpn, Meox1]         # explicit list
# default_genes: "Cd36, Pdpn, Meox1"       # comma-separated string
# defaults:                                 # nested form, matches the
#   genes: [Cd36, Pdpn, Meox1]              #   bundle's defaults/genes
#                                           #   HDF5 path
# preGenes: "Cd36, Pdpn, Meox1"            # legacy alias
```

## Field-by-field {-}

| Key | Required | Default | Notes |
| --- | --- | --- | --- |
| `output`            | yes  | —             | Root directory for the per-study subfolder. The bundle ends up at `<output>/<studyID>/<studyID>.scminer.h5`. |
| `study.ID`          | yes  | —             | Numeric or string. Used as the subfolder name. |
| `study.studyAbbr`   | yes  | —             | Short identifier (URL slug). |
| `study.longTitle`   | yes  | —             | Full study description. |
| `study.shortTitle`  | yes  | —             | One-line label. |
| `species`           | no   | `""`          | e.g. `"Mus musculus"`. Also auto-filled from the manifests if missing. |
| `coordinate`        | no   | `"UMAP"`      | Both the meta field and the pData column prefix — `<coord>_1` and `<coord>_2`. |
| `cellID`            | no   | `"cellID"`    | pData column carrying the cellID. Rownames are also written into this column. |
| `cellType`          | no   | `"cellGroup"` | pData column carrying the cell-type label. |
| `cellGroup`         | no   | `cellType`    | pData column carrying the cell-group label. |
| `geneSymbol`        | no   | `"geneSymbol"`| fData column carrying the gene symbol. |
| `cluster_palette`   | no   | `"npg"`       | ggsci palette name for auto-generated cluster colors. One of `npg, aaas, lancet, nejm, jama, jco, ucscgb, d3, locuszoom, igv, uchicago, startrek, tron, futurama, rickandmorty, simpsons` (see the [ggsci vignette](https://cran.r-project.org/web/packages/ggsci/vignettes/ggsci.html)). Falls back to a baked-in NPG palette if `ggsci` isn't installed. |
| `input.expression`  | yes  | —             | Path to the expression ExpressionSet RDS. |
| `input.activity`    | no   | none          | Path to the activity ExpressionSet RDS. Rows ending in `_TF` / `.TF` go to `activity_tf`; `_SIG` / `.SIG` go to `activity_sig`. |
| `input.networks`    | no   | none          | Path to the networks TSV. |
| `default_genes`     | no   | none          | Genes the Shiny app pre-selects on launch. Accepts a YAML list (`[A, B, C]`), a comma- / semicolon- / newline-separated string, the nested form `defaults: { genes: [...] }`, or the legacy alias `preGenes`. Parsed via `parse_default_genes()`; any gene not in the bundle's master index is dropped with a warning, so the on-disk `defaults/genes` is always referentially valid. |

## Parsing the config without loading the RDS files {-}

`load_study_config()` parses the YAML, validates required keys, and
fills in defaults. It does **not** touch any RDS / TSV files referenced
by `input.*` — useful for sanity-checking the config or building a
custom pipeline:

```{r parse-config, eval=FALSE}
cfg <- scminerViewer::load_study_config("config.yml")
str(cfg)
```

See Chapter \@ref(r-api) for the full list of staged helpers
(`load_study_config`, `extract_cells`, `extract_genes`,
`extract_expression`, `extract_activity`, `read_networks`,
`fill_clusters`) that `prepare_study()` is composed of.

## Auto-generated cluster metadata {-}

`prepare_study()` always emits a `study_meta/<studyID>_study_meta.csv`
with full color + label-position info, even when your input
ExpressionSet only carries `cellType` (no precomputed cluster table).
Two pieces are auto-filled by `fill_clusters()`:

* **`Color`** — `n_clusters` colours pulled from the configured ggsci
  palette (`cluster_palette` above). Default is NPG (Nature Publishing
  Group). If `ggsci` isn't installed, a baked-in 10-colour NPG
  fallback is used.
* **`Label_1`, `Label_2`** — per-cluster centroids: `mean(coord1)` and
  `mean(coord2)` over the cells in each cluster. The cluster plot uses
  these to place the cluster-name annotations.

Existing values in the clusters data.frame (when one is supplied) are
**preserved**; only NA / empty cells are filled in. So you can hand-pick
a few colours and let the rest fall back to ggsci.

## Pre-selecting genes on app launch (`default_genes`) {-}

`prepare_study()` can pre-populate the Shiny app's gene picker with a
small, curated set of "canonical" genes for that study — the same
mechanism the scMINER Portal calls `preGenes`. Declare the list in
your YAML config under any of these keys (they all map to the same
internal value):

```yaml
default_genes: [Cd36, Pdpn, Meox1]              # explicit list (recommended)

default_genes: "Cd36, Pdpn, Meox1"              # comma-separated string

defaults:                                        # nested form (mirrors the
  genes: [Cd36, Pdpn, Meox1]                    #   on-disk HDF5 group)

preGenes: "Cd36, Pdpn, Meox1"                   # legacy alias from the
                                                 #   scMINER Portal JSON export
```

`load_study_config()` normalises any of those into a deduplicated
character vector at `cfg$default_genes`. Internally the helper used is
`parse_default_genes()`:

```{r parse-default-genes, eval=FALSE}
scminerViewer::parse_default_genes("Cd36, Pdpn; Meox1\nCd36")
#> [1] "Cd36"  "Pdpn"  "Meox1"
```

When you run `prepare_study(config_path)`, the validated subset of
default genes (only the ones that exist in the bundle's master gene
index) is written to the HDF5 group `defaults/genes`. The R Shiny app
and the Python webui both auto-select those genes on launch:

```r
study <- load_study("study/2327/2327.scminer.h5")
study$default_genes
#> [1] "Cd36" "Pdpn" "Meox1"
```

You can also set default genes **programmatically** from a Biobase
pipeline (no YAML required):

```{r prepare-default-genes-eset, eval=FALSE}
prepare_study_from_eset(
  out_dir         = "study",
  expression_eset = eset,
  meta            = list(studyID = "2327", studyAbbr = "Tex",
                          longTitle = "T-cell exhaustion (Tex)",
                          shortTitle = "Tex"),
  default_genes   = c("Cd36", "Pdpn", "Meox1")
)
```

`validate_default_genes()` filters the list against the bundle's
master gene index before write — any unknown symbol is dropped with a
warning naming the offenders, so the on-disk `defaults/genes` is
always referentially valid:

```{r validate-default-genes, eval=FALSE}
scminerViewer::validate_default_genes(
  c("Cd36", "Pdpn", "MissingGene"),
  master_genes = c("Cd36", "Pdpn", "Meox1")
)
#> Warning: default_genes: 1/3 not in master gene list (dropping): MissingGene
#> [1] "Cd36" "Pdpn"
```
