# Config YAML reference {#config-yaml}

`prepare_study(config_path)` is the YAML-driven entry point for the
full pipeline. It points at a scMINER `ExpressionSet` RDS (and
optional activity + networks) and writes both the graph-import layout
and the lazy-mode bundle in one call.

Two examples ship with the project:

| File | Role |
| --- | --- |
| `scminerViewer/inst/extdata/example_config.yml` | Annotated **template** — every field documented inline; reachable via `system.file()`. |
| `data/example/2327.yml`                          | Concrete **runnable** config keyed to the actual 2327 study (Tex). Edit `input.*` paths and run. |

Copy the template:

```{r copy-template, eval=FALSE}
file.copy(
  system.file("extdata", "example_config.yml", package = "scminerViewer"),
  "config.yml"
)
```

## Full schema {-}

```yaml
# Where to write everything — the bundle lands at
# <output>/<studyID>/<studyID>.scminer.h5 (per-study subfolder).
output: data

# Study identity
study:
  ID:         "2327"
  studyAbbr:  tex
  longTitle:  >
    Single-cell RNA sequencing of progenitor, intermediate and
    terminally exhausted CD8+ T cells from tumors and chronic viral
    infection
  shortTitle: Tex

# Sample-level metadata
species:    Mus musculus
coordinate: UMAP            # also the pData column prefix (UMAP_1, UMAP_2)

# pData / fData column names in the source ExpressionSet
cellID:     cellID          # rownames are also assigned to this column
cellType:   cellGroup       # the cell-type / cluster label
cellGroup:  cellGroup       # falls back to cellType when omitted
geneSymbol: geneSymbol      # column in fData carrying the gene symbol

# Input files
input:
  expression: inputs/2327_expression.rds   # required, Biobase ExpressionSet
  activity:   inputs/2327_activity.rds     # optional, ExpressionSet
                                            # rows ending in _TF / _SIG split
                                            # into the two activity matrices
  networks:   inputs/2327_networks.tsv     # optional, tab-separated:
                                            # source, target, NetworkType,
                                            # CellGroup, mi, pearson,
                                            # spearman, rho, pvalue
```

## Field-by-field {-}

| Key | Required | Default | Notes |
| --- | --- | --- | --- |
| `output`            | yes  | —             | Root directory for the per-study subfolder. The bundle ends up at `<output>/<studyID>/<studyID>.scminer.h5`. |
| `study.ID`          | yes  | —             | Numeric or string. Used as the subfolder name. |
| `study.studyAbbr`   | yes  | —             | Short identifier (URL slug). |
| `study.longTitle`   | yes  | —             | Full study description. |
| `study.shortTitle`  | yes  | —             | One-line label. |
| `species`           | no   | `""`          | e.g. `"Mus musculus"`. Also auto-filled from the manifests if missing. |
| `coordinate`        | no   | `"UMAP"`      | Both the meta field and the pData column prefix — `<coord>_1` and `<coord>_2`. |
| `cellID`            | no   | `"cellID"`    | pData column carrying the cellID. Rownames are also written into this column. |
| `cellType`          | no   | `"cellGroup"` | pData column carrying the cell-type label. |
| `cellGroup`         | no   | `cellType`    | pData column carrying the cell-group label. |
| `geneSymbol`        | no   | `"geneSymbol"`| fData column carrying the gene symbol. |
| `input.expression`  | yes  | —             | Path to the expression ExpressionSet RDS. |
| `input.activity`    | no   | none          | Path to the activity ExpressionSet RDS. Rows ending in `_TF` / `.TF` go to `activity_tf`; `_SIG` / `.SIG` go to `activity_sig`. |
| `input.networks`    | no   | none          | Path to the networks TSV. |

## Parsing the config without loading the RDS files {-}

`load_study_config()` parses the YAML, validates required keys, and
fills in defaults. It does **not** touch any RDS / TSV files referenced
by `input.*` — useful for sanity-checking the config or building a
custom pipeline:

```{r parse-config, eval=FALSE}
cfg <- scminerViewer::load_study_config("config.yml")
str(cfg)
```

See Chapter \@ref(r-api) for the full list of staged helpers
(`load_study_config`, `extract_cells`, `extract_genes`,
`extract_expression`, `extract_activity`, `read_networks`) that
`prepare_study()` is composed of.
