# Bundle format & shard tree {#bundle-and-shards}

The `.scminer.h5` bundle is the R↔Python contract. It's a small HDF5
file (~80 MB for an 8K-cell study) that contains everything the apps
need at startup — except the matrix values, which stay on disk in the
shard tree and are read lazily.

## HDF5 bundle layout (`bundleVersion = 1`) {-}

```
/meta/
  studyID, studyAbbr, longTitle, shortTitle, species, coordinate  (string)
  bundleVersion                                                   (int, = 1)

/cells/
  cellID    (N,)   string
  cellType  (N,)   string
  cellGroup (N,)   string
  coord1    (N,)   float64
  coord2    (N,)   float64

/clusters/
  cellType  (K,)   string
  count     (K,)   int32
  color     (K,)   string  (hex)
  label_1   (K,)   float64    optional — x for cluster label placement
  label_2   (K,)   float64    optional — y for cluster label placement

/genes/
  symbol    (G,)   string     master picker list

/index/                       optional, "which genes are available"
  expression   (G_e,) string  present iff expression shards exist
  activity_tf  (G_t,) string
  activity_sig (G_s,) string

/defaults/                    optional, "load these on app startup"
  genes        (D,)   string

/network_tf/                  optional
  source, target, cellType                       string
  mi, pearson, spearman, rho, pvalue             float64
/network_sig/                 optional, same columns
```

Strings are UTF-8 variable-length. Scalar string datasets are stored
as 1-element arrays by the R writer; the Python reader unwraps them
automatically.

**Compatibility rules** (both packages MUST honour):

* Unknown groups → treat as future-bundle-versions; ignore.
* Any optional group may be absent.
* Scalar string datasets may be either 1-element arrays or true
  scalars.

## Shard tree {-}

Sharded by the first lowercase letter of each gene name (or `nm` for
non-alphabetic), with a single cell-ID header per matrix type:

```
<shard_dir>/
├── expression_files/<studyID>/
│   ├── meta.csv                  one line, comma-separated cellIDs
│   └── <letter>/<gene>.csv.gz    gzipped 1-row CSV of N values
└── activity_files/<studyID>/
    ├── meta.csv                  shared by TF and SIG (may be a cell subset)
    ├── TF/<letter>/<gene>.csv.gz
    └── SIG/<letter>/<gene>.csv.gz
```

* `<letter>` is `tolower(substr(gene, 1, 1))` if alphabetic, else `nm`.
* Path is derived from the gene name, not the manifest's `File` column.
* Cell-ID column order is read once per matrix type from `meta.csv`;
  a permutation index aligns shard columns to `cells$cellID`, so
  missing cells become zero columns automatically.
* The activity `meta.csv` is **shared** between TF and SIG and may
  list a subset of the cells in the expression meta (activity is
  computed only for cells passing some upstream filter).

## Lazy contract {-}

`load_study(bundle_path)` reads the bundle and returns a thin handle
that knows where the shards are. No matrix values are loaded.
`gene_values(study, gene, relationship)` (R) /
`study.gene_values(gene, relationship)` (Python) reads exactly one
shard from disk, caches the aligned vector on the study handle, and
returns it.

```{r lazy-r, eval=FALSE}
study <- scminerViewer::load_study("data/2327/2327.scminer.h5")
vals  <- scminerViewer::gene_values(study, "Mrpl15", "Express_normalized")
length(vals)   # nrow(study$cells)
```

```python
import scminer_viewer
study = scminer_viewer.load_study("data/2327/2327.scminer.h5")
vals  = study.gene_values("Mrpl15", "Express_normalized")
vals.shape  # (n_cells,)
```

## Discovering studies under a multi-study root {-}

`discover_studies(root_dir)` scans for `<studyID>/<studyID>.scminer.h5`
patterns and returns a row-per-study data frame. Useful for scripting
or as the data source for the browser's card-grid index page.

```{r discover-r, eval=FALSE}
df <- scminerViewer::discover_studies("data")
nrow(df)   # number of studies the browser will list
```

```python
import scminer_viewer
df = scminer_viewer.discover_studies("data")
df.shape[0]
```
