# Troubleshooting {#troubleshooting}

## "lazy-load database is corrupt" / "interrupted promise evaluation" {-}

R memory-maps the package's `.rdb` file during `library()`. If a
second R process reinstalls the package while the first one still has
it loaded, you'll see one of:

```
Error: lazy-load database '.../scminerViewer.rdb' is corrupt
Warning: restarting interrupted promise evaluation
Error in R_decompress1
```

**Fix** — always the same recipe:

1. Quit **every** R session that has the package loaded (in RStudio:
   **Session → Quit Session**, `Cmd/Ctrl + Q`; restarting alone isn't
   enough — the OS still has the old `.rdb` mapped until the process
   exits).
2. Wipe the install directory:

   ```sh
   rm -rf "$HOME/StJude/Research/R_lib/scMINERViewer/scminerViewer"
   ```
3. Reinstall from a fresh R session.

To avoid it: always restart R **before** reinstalling
(`Cmd/Ctrl + Shift + F10` in RStudio).

## `meta.csv not found: data/<sid>/expression_files/<sid>/meta.csv` {-}

The bundle resolved its `shard_dir` to its parent directory but the
shards live somewhere else. Either move the shards into the bundle's
parent:

```sh
mv data/example/expression_files data/2327/
mv data/example/activity_files   data/2327/
```

…or tell the app where they are at launch time:

```{r shard-dir-r, eval=FALSE}
run_app("data/2327/2327.scminer.h5",
        shard_dir = "data/example", port = 8000)

run_browser("data", shard_dir = "data/example", port = 8000)
```

```sh
scminer-viewer run data/2327/2327.scminer.h5 --shard-dir data/example
scminer-viewer browse data --shard-dir data/example
```

## `No documentation for 'prepare_study'` {-}

The package was installed without man pages. Regenerate via roxygen2:

```{r roxy-rebuild, eval=FALSE}
roxygen2::roxygenise("scminerViewer")    # writes NAMESPACE + man/*.Rd
```

…then reinstall.

## `could not find function "%>%"` {-}

Older versions of the package used magrittr's pipe in plot helpers
without importing `magrittr`. The current version uses base R 4.1+
`|>` throughout, so reinstall to get the fix.

## Plotly "No trace type specified … Setting the mode to markers" {-}

These warnings used to come from `plotly::plotly_empty()` placeholders.
The current version uses an explicit empty scatter trace
(`.empty_plot()`) and the warnings should be gone — if they still
appear, reinstall.

## `run_browser` shows no studies {-}

Run `discover_studies(root)` from R (or `scminer-viewer list <root>`
in Python) to see what the browser would find. Each study must:

* be in its own subfolder named exactly the studyID, and
* contain a file named `<studyID>.scminer.h5` (matching the folder name).

A study folder produced before the multi-study refactor sits **flat**
under the root (`<root>/Cell/`, `<root>/Gene/`, etc.); it must be
wrapped in `<root>/<studyID>/` to be picked up. The
`build_2327_bundle.R` script does this automatically.

## A gene I expect is missing from the picker {-}

The picker is populated from the bundle's `genes/symbol` group, which
is the **master** gene list (every gene in the study, regardless of
which matrices have shards). If a gene is missing here, regenerate
the bundle.

If a gene is in the picker but `gene_values()` returns NULL for a
particular relationship, the gene isn't in that matrix's index —
check `study$expression_index` / `$activity_tf_index` /
`$activity_sig_index` (R) or `study.expression_index` etc. (Python).
Activity matrices typically have a smaller index than expression
(only TFs / SIGs are computed).

## Bundle is huge (>1 GB) {-}

`bundleVersion == 2` stores **indexes only** — no matrix values — so
a healthy bundle is tens of MB even for studies with hundreds of
thousands of cells. If your bundle is >1 GB, you're probably looking
at a stale `bundleVersion == 1` artifact from before the lazy-mode
refactor. Rebuild via `prepare_study()` or the migration recipe in
Chapter \@ref(tutorial).

## A specific cluster disappears with sampling on {-}

The per-cluster proportional sampling keeps at least one cell per
cluster (`max(1L, round(n * pct / 100))`) so no cluster vanishes
entirely. If you see an empty cluster, the cluster has zero cells in
the underlying study — check `study$clusters$count` (R) /
`study.clusters['count']` (Python).
