# Troubleshooting {#troubleshooting}

## "lazy-load database is corrupt" / "interrupted promise evaluation" {-}

R memory-maps the package's `.rdb` file during `library()` and
**doesn't release the handle on reinstall**. If a second R process
reinstalls the package while the first one still has it loaded — or
the same session reinstalls itself — you'll see one of:

```
Error: lazy-load database '.../scminerViewer.rdb' is corrupt
Warning: restarting interrupted promise evaluation
Error in R_decompress1
```

The crude fix is the recipe most blogs prescribe: quit every R session
holding the package, wipe the install dir, reinstall from a fresh
session. Three better options:

### 1. Don't `library()` during development — use `devtools::load_all()` {-}

`load_all()` reads the package source directly; no `.rdb` is ever
written, so the corruption literally cannot happen.

```{r load-all, eval=FALSE}
devtools::load_all("scminerViewer")     # Cmd/Ctrl + Shift + L in RStudio
```

This is what 90% of R package developers use day-to-day. `library()`
is for end-users.

### 2. RStudio: tick "Restart R prior to install" {-}

**Tools → Project Options… → Build Tools → Restart R prior to install
and clean** — then **Build → Install Package** (`Cmd/Ctrl + Shift + B`)
kills the R session before each install. Costs you the session state,
but guarantees a clean `.rdb`.

### 3. Use the shipped safe-install helper for scripted installs {-}

`scminerViewer/inst/scripts/install.R` unloads the namespace,
GCs to release mmap pages, wipes the install dir, then installs in a
fresh `callr` subprocess — so the parent R never holds a stale
handle.

```{r safe-install-r, eval=FALSE}
source(system.file("scripts", "install.R", package = "scminerViewer"))
safe_install("scminerViewer")                # auto-detects .libPaths()[1]
safe_install("scminerViewer", lib = "/path/to/lib")
```

Or from a shell:

```sh
Rscript scminerViewer/inst/scripts/install.R scminerViewer
```

The helper cannot unload the package from *other* R sessions on the
same machine — those have to quit on their own first.

### Last resort: manual cleanup {-}

If everything's stuck:

1. **Quit every R session** that has the package loaded — in RStudio
   use **Session → Quit Session** / `Cmd/Ctrl + Q`. (Restarting alone
   isn't enough; the OS still has the old `.rdb` mmap'd until the
   process *exits*.)
2. Wipe the install directory from a shell:

   ```sh
   rm -rf "$HOME/StJude/Research/R_lib/scMINERViewer/scminerViewer"
   ```
3. Reinstall from a fresh R session.

## `meta.csv not found: data/<sid>/expression_files/<sid>/meta.csv` {-}

The bundle resolved its `shard_dir` to its parent directory but the
shards live somewhere else. Either move the shards into the bundle's
parent:

```sh
mv data/example/expression_files data/2327/
mv data/example/activity_files   data/2327/
```

…or tell the app where they are at launch time:

```{r shard-dir-r, eval=FALSE}
run_app("data/2327/2327.scminer.h5",
        shard_dir = "data/example", port = 8000)

run_browser("data", shard_dir = "data/example", port = 8000)
```

```sh
scminer-viewer run data/2327/2327.scminer.h5 --shard-dir data/example
scminer-viewer browse data --shard-dir data/example
```

## `No documentation for 'prepare_study'` {-}

The package was installed without man pages. Regenerate via roxygen2:

```{r roxy-rebuild, eval=FALSE}
roxygen2::roxygenise("scminerViewer")    # writes NAMESPACE + man/*.Rd
```

…then reinstall.

## `could not find function "%>%"` {-}

Older versions of the package used magrittr's pipe in plot helpers
without importing `magrittr`. The current version uses base R 4.1+
`|>` throughout, so reinstall to get the fix.

## Plotly "No trace type specified … Setting the mode to markers" {-}

These warnings used to come from `plotly::plotly_empty()` placeholders.
The current version uses an explicit empty scatter trace
(`.empty_plot()`) and the warnings should be gone — if they still
appear, reinstall.

## `run_browser` shows no studies {-}

Run `discover_studies(root)` from R (or `scminer-viewer list <root>`
in Python) to see what the browser would find. Each study must:

* be in its own subfolder named exactly the studyID, and
* contain a file named `<studyID>.scminer.h5` (matching the folder name).

A study folder produced before the multi-study refactor sits **flat**
under the root (`<root>/Cell/`, `<root>/Gene/`, etc.); it must be
wrapped in `<root>/<studyID>/` to be picked up. The
`build_2327_bundle.R` script does this automatically.

## A gene I expect is missing from the picker {-}

The picker is populated from the bundle's `genes/symbol` group, which
is the **master** gene list (every gene in the study, regardless of
which matrices have shards). If a gene is missing here, regenerate
the bundle.

If a gene is in the picker but `gene_values()` returns NULL for a
particular relationship, the gene isn't in that matrix's index —
check `study$expression_index` / `$activity_tf_index` /
`$activity_sig_index` (R) or `study.expression_index` etc. (Python).
Activity matrices typically have a smaller index than expression
(only TFs / SIGs are computed).

## Bundle is huge (>1 GB) {-}

`bundleVersion == 2` stores **indexes only** — no matrix values — so
a healthy bundle is tens of MB even for studies with hundreds of
thousands of cells. If your bundle is >1 GB, you're probably looking
at a stale `bundleVersion == 1` artifact from before the lazy-mode
refactor. Rebuild via `prepare_study()` or the migration recipe in
Chapter \@ref(tutorial).

## A specific cluster disappears with sampling on {-}

The per-cluster proportional sampling keeps at least one cell per
cluster (`max(1L, round(n * pct / 100))`) so no cluster vanishes
entirely. If you see an empty cluster, the cluster has zero cells in
the underlying study — check `study$clusters$count` (R) /
`study.clusters['count']` (Python).
